---
title: "Yearly Bike Sharing Data for Year 2020"
author: "Steven Pyae"
date: "5/12/2021"
output: pdf_document
---
## Data packages Credited to Divvybikes under this [license](https://www.divvybikes.com/data-license-agreement)

### Packages Used
```{r loading Environment}
library(tidyverse)
library(lubridate)
library(ggplot2)
setwd("D:/R/Data/Google Course Data/CSV Files")
```


 * Divvy_Trips_2020_Q1.csv - q1_2020
 * 202005-divvy-tripdata.csv - may_2020
 * 202006-divvy-tripdata.csv - june_2020
 * 202007-divvy-tripdata.csv - july_2020
 * 202008-divvy-tripdata.csv - aug_2020
 * 202009-divvy-tripdata.csv - sep_2020
 * 202010-divvy-tripdata.csv - oct_2020
 * 202011-divvy-tripdata.csv - nov_2020
 * 202012-divvy-tripdata.csv - dec_2020
 
### Loading Data sets
```{r loading dataset}
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
may_2020 <- read_csv("202005-divvy-tripdata.csv")
june_2020 <- read_csv("202006-divvy-tripdata.csv")
july_2020<- read_csv("202007-divvy-tripdata.csv")
aug_2020<- read_csv("202008-divvy-tripdata.csv")
sep_2020<- read_csv("202009-divvy-tripdata.csv")
oct_2020<- read_csv("202010-divvy-tripdata.csv")
nov_2020 <- read_csv("202011-divvy-tripdata.csv")
dec_2020<- read_csv("202012-divvy-tripdata.csv")
```


### July_2020's Dates are formatted into POSIXct to bind with other datasets
```{r formatting dates}
july_2020 <- mutate(july_2020, started_at =as.POSIXct(format(strptime(started_at, "%d/%m/%Y %H:%M"), "%Y-%m-%d %H:%M")))
july_2020 <- mutate(july_2020, ended_at =as.POSIXct(format(strptime(ended_at, "%d/%m/%Y %H:%M"), "%Y-%m-%d %H:%M")))
```
### Binding all the datasets into one
```{r binding data sets}
all_trips <- bind_rows(q1_2020,may_2020,june_2020,july_2020,aug_2020,sep_2020,oct_2020,nov_2020)
```
### Data Cleaning
```{r cleaning}
# Remove the lat, lng, lat, lng
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))
# check for consistent name conventions
table(all_trips$member_casual)

#add columns that list the date,month, day and year of each ride
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```
### Adding additional columns: ride_length, day_of_week
```{r ride_length and day_of_week}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
# Convert "ride_length" from Factor to numeric so we can run calculations on the data
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

### Data contains negative ride_length and non-applicable numbers, Filtering
```{r filtering}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
# Remove non existent ride lengths
all_trips_v2 <- all_trips_v2[!is.na(all_trips_v2$ride_length),]
```

## Plots

Analysing the results

```{r No of rides vs rider type}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title = "Number of Rides by Rider Type")

```
```{r Avg duration vs rider type}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title = "Average Duration by Rider Type")

```

